SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2013-03-22 paper/paper.tex
paper: major updates.

\title{SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization\\$\;$\\DRAFT}
%\subtitle{Subtitle Text, if any}

@@

Defining an efficient and effective shrink method requires a good understanding
of how shrinking in QuickCheck works and the semantics of the property and
program being evaluated.  Bad definitions can be so inefficient that they are
unusable.

In addition, shrinking is one side of the coin when it comes to making counterexamples more
understandable.  The other side is generalization.  For example, the following
is a relatively small counterexample returned when QuickChecking an (incorrect)
implementation in Xmonad, a popular X11 window manager written in
Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
Functional programmers should be familiar with having to debug a ``wall of
text'' when showing data structures.  What if instead a counterexample like the
following were returned by the testing framework:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the counterexample, meaning that for any well-typed values that
replace \ttp{x0}, \ttp{x1}, \ttp{x2}, and \ttp{x3}, the counterexample still
holds.

@@

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  More importantly, the
time required to shrink scales linearly with the size of the
counterexample---other common shrink implementations often scale in polynomial
time or worse.  And the counterexamples returned are two to five times smaller.

@@

prop :: T -> Property
prop t = pre t ==> post t
\end{code}
  \caption{QuickCheck and SmallCheck for a product-type input.}
  \label{fig:initial}
\end{figure}

@@

A programmer might try to control the complexity by truncating lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, we might redefine shrink as follows:
%
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space; the maximum number of possible new values is $10^5$ in this case.
While truncation controls the blowup, the downside is that potentially smaller
counterexamples may be omitted.

@@

The programmer might ask herself: ``What if we omit the need for shrinking
counterexamples altogether?''  SmallCheck is another testing framework for
Haskell that does just this.  SmallCheck is guaranteed to return a smallest
counterexample, if one exists~\cite{sc}.  SmallCheck does this by enumerating
all possible inputs, ordered from smallest to largest, up to some user-defined
bound.  While SmallCheck is effective for testing many programs and properties
(in accordance with the \emph{small scope hypothesis}~\cite{jackson}),
counterexamples to even relatively simple properties may be infeasible to
discover due to input-space explosion.

@@

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, SmallCheck
fails long before it reaches a depth required to discover a counterexample.  (A
related library named Feat combines some aspects of SmallCheck and
QuickCheck~\cite{feat}; we discuss it in Section~\ref{sec:related}).

@@

Here we provide some illustrative comparisons between SmartCheck's shrinking
against other shrinking approaches.  As a representative example, we use the
example presented in Section~\ref{sec:intro}.  The results for SmartCheck are
fairly generalizable to other types and properties since SmartCheck's reduction
algorithm has no dependencies on the specifics of either; however, user-defined
shrink methods may vary significantly based on the property and counterexample
type.

@@

To motivate the benefits of SmartCheck, consider the graphs in
Figure~\ref{fig:graphs} and the Table~\ref{table:results}, giving the results
for using QuickCheck and SmartCheck on the program shown in
Figure~\ref{fig:initial} (we omit SmallCheck, since as noted, it is not even
feasible for the example).  We graph counterexample size and execution time.
Out of 2000 runs, we show the number of runs resulting in a shrunk
counterexample of a given size (by counting \ttp{N} constructors), and we show
the number of runs executing within a given time limit (in seconds).  Note that
we provide two execution-time graphs, since the scale of time scales differ
dramatically.  In the table, we show the mean, median, median, and the results
at the 95th percentile values.  (The distributions produced by these values are
not necessarily Gaussian, so we omit standard the deviation.)

@@

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck; thus the time required by SmartCheck to shrink a counterexample
is on average just 0.2 seconds (0.10 - 0.08).  As can be seen in the graphs,
SmartCheck's execution time scales linearly (like QuickCheck, without
shrinking).  Although it is difficult to tell from the graphs, truncation and
tuple shrinking produce very long tails, with some counterexamples taking over a
minute to shrink.  SmartCheck is largely insensitive to the size of the original
counterexample provided.

@@

We focus on algebraic data types represented as ``sums of products''.  We
sometimes refer to the elements of a sum type as a \emph{variant} that is
\emph{tagged} by a constructor.  For example, the type
%
\begin{code}
Maybe a = Just a | Nothing
\end{code}
%
\noindent
contains two variants, tagged by the constructors \ttp{Just} and \ttp{Nothing},
respectively.

To present the algorithms in the following sections, we provide some definitions
in the form of methods of a \ttp{SubTypes} type class.  SmartCheck requires
instances for data being analyzed.  In Section~\ref{sec:implementation}, we
describe how instances for the type class are derived automatically.
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size     :: a -> Size
  index    :: a -> Idx -> Maybe SubVal
  replace  :: a -> Idx -> SubVal -> a
  constr   :: a -> String
  constrs  :: a -> [String]
  baseType :: a -> Bool
  subVals  :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{baseType} is true when the type of the value is an ``uninteresting
  type'', which generally means it is user-defined and has positive arity.
  We discuss the purpose of base types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non base-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

baseType (3 :: Int) = True
baseType tree       = False
baseType L          = False

@@

%% Finally, \ttp{baseType} is true for our tree type but false for a
%% structureless type, such as \ttp{Int}.

@@

The approach taken by typical \ttp{shrink} implementations is the most naive
graph search algorithm: it simply generates a random set of potential vertexes
to check in the graph.  However, it does not exploit that the original
counterexample \ttp{cex} is already in a neighborhood of other likely
counterexamples.  So rather than choose random vertexes, it may be more
efficient to test other values that are closely related to \ttp{cex}.  By
restricting the search to \ttp{cex}'s neighborhood, a smaller counterexample
might be missed, and only a local minimum is returned.  However, with a large
set of values, finding a global minimum by random sampling vertexes in the graph
is possible but too inefficient, as we saw from the example in the introduction.

\subsection{Reduction Algorithm Overview}\label{sec:reduct}
The algorithm we present for efficiently searching for new counterexamples is an
instance of greedy breadth-first search over a tree structures, representing a
value.  At each node, during the traversal, we generate arbitrary
\emph{structurally} smaller sub-values and build a new value from that, leaving
the remainder of the tree unchanged.  Intuitively, by a \emph{structurally
  smaller value}, we mean one with fewer constructors.  We continue until we
reach a fixed-point.

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  At the beginning of execution, we
assume that a counterexample has been obtained somehow; our implementation
relies on QuickCheck to generate an initial counterexample and to generate
arbitrary well-typed random values (by the \ttp{arbitrary} method).  The
reduction begins at the first proper sub-value of \ttp{cex}; call it \ttp{v}.
When the index \ttp{idx} becomes out-of-bounds and returns \ttp{Nothing}, the
algorithm terminates.  Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}.  Recall that the algorithm begins at
the \emph{first} sub-value of the counterexample rather than the 0th sub-value.
The 0th sub-value of \ttp{cex} is \ttp{cex} itself.  No invariant of the
algorithm would be violated by beginning with the 0th sub-value, and in
particular, the algorithm would still terminate.  However, we begin with a
strict sub-value because of the hypothesis.  The hypothesis is that in the
search space of possible values from a given type \ttp{T}, starting with a
particular counterexample \ttp{cex}, we are more likely to find other
counterexamples from the same variant as \ttp{cex}.  For example, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

@@

\subsubsection{Base Types}\label{sec:base}

The number of constructors in an algebraic data value primarily determines how
large it is.  In our experience, the size of fields that come from primitive
types, like \ttp{Int} or \ttp{Char}, have less of a role in understanding a
counterexample, so in general, we want to prevent our reduction algorithm from
spending time shrinking values from these types.

Consequently, we distinguish a class of types referred to as \emph{base types}.
If the reduction algorithm encounters a base type, it is ignored.  By default,
this includes numeric types (e.g., \ttp{Int}, \ttp{Word}, \ttp{Float}, etc.,
\ttp{Bool}, and \ttp{Char}.  We also treat recursive container types like lists
specially: lists of base types are considered to be base types themselves.  This
prevents the algorithm from analyzing, for example, strings.  A list of
non-base-types, however, is shrunk by the algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a base type.  Doing so effectively treats values from that type as
``black boxes'' for SmartCheck.  Doing so can help make SmartCheck more
efficient if the user knows that some portion of the structure cannot be shrunk
or is irrelevant to the property.

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard in which a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
is true if in \ttp{cex}, the value at index \ttp{idx0}, is a child of index
\ttp{idx1}, in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  This
prevents the algorithm from trying to generalize sub-values that are abstracted
away already since their parents have been generalized.

@@

\paragraph{Soundness}
The extrapolation algorithm is unsound in two ways.  First, it extrapolates from
a set of counterexamples to a universal claim, similar to QuickSpec~\cite{qs}.
By tuning the parameters, the risk of an unsound generalization is reduced by
requiring more or larger values to fail the property.

Second, note that each index is generalized with respect to the concrete
sub-values that exist in the counterexample.  So in some cases, a formula may be
returned that is overly general.  For example, consider the counterexample in
which both fields of \ttp{Add} contain values causing the failure:
%
\begin{code}
Add (Div (C 1) (C (-2)) (C 2))
    (Div (C 0) (C (-1)) (C 1))
\end{code}
%
\noindent
since no matter what random value the first field of \ttp{Add} is replaced with,
the property fails by the second field and vice versa for the second.  The
universal generalization algorithm might return the formula
\noindent
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
The algorithm could be modified so that when attempting to generalize a
sub-value at an index, all previously-generalized indexes are re-generalized
simultaneously, by taking the cross-product of random values for the current
index and for the previously-generalized indexes.  While this reduces the
potential for over-generalizing a counterexample, it is exponential in the
number of indexes generalized.  For universally quantified formulas, the reader
should read it as shorthand for quantifying each variable independently and
taking the conjunction of formulas.  For example, instead of
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
one should read
%
%
\begin{code}
      forall values x0 .
        Add x0 (Div (C 0) (C (-1)) (C 1))
  and forall values x1 .
        Add (Div (C 1) (C (-2)) (C 2)) x1
\end{code}

@@

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there a counterexample
containing every possible variant as a sub-value, then we abstract it.  For
example, the following are all counterexamples:
%
\begin{code}
Div (C 7) (Add (C (-5)) (C 5))
Div (Add (C 3) (C 2)) (C 0)
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors c .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (c\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

Figure~\ref{fig:cex-gen} illustrates a property of the form (\ttp{pre ==>
  post}).  Points are specific counterexamples

 the situation in which we have some values
and a property in the case that some values fail the property of the form
.  Our goal is to find additional counterexamples in the
shaded region, in which \ttp{pre} is satisfied, \ttp{post} is not satisfied, and
the counterexample is not already characterized by a universal sub-value
generalization formula.  Iterating, we increasingly cover the shaded region
with new formulas until no more uncovered counterexamples exist or the testing
tool fails to find new counterexamples.

@@

Figure~\ref{fig:matches} shows the shape-matching algorithm used for precondition
strengthening.  The function takes a candidate counterexample and a list of
previous counterexamples, together with the indexes at which they have been
generalized.  The basic idea of the algorithm is to determine whether two values
have the same ``shape''.  Intuitively, we consider two values to have the same
shape if in their respective tree representations, their constructors at each
node in the tree match, ignoring all base types (Section~\ref{sec:base}).
That is, intuitively, for values \ttp{a} and \ttp{b},
%
\begin{code}
subVals a == subVals b
\end{code}
%
\noindent
Furthermore, indexes that have been universally generalized (and all their
children) match any value, since a universally generalized index indicates that
counterexamples have been found for any possible constructor.  In our
implementation, we also consider existentially generalized indexes to match any
value.  Doing so is a design choice that is more aggressive about covering the
space of counterexamples at the risk of omitting some.

>>>>

\title{SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization}
%\subtitle{Subtitle Text, if any}

@@

Defining an efficient and effective shrink method requires a good understanding
of how shrinking in QuickCheck works and the semantics of the property and
program being evaluated.  Bad definitions can be so slow or so ineffective at
shrinking that they are unusable.

In addition, shrinking is one side of the coin when it comes to making
counterexamples more understandable: the other side is generalization.  For
example, the following is a relatively small counterexample returned when
QuickChecking a property in Xmonad (with a bug inserted), a popular X11 window
manager written in Haskell~\cite{xmonad}:
%
\medskip%
\begin{sloppypar}
%\footnotesize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
(This counterexample uses Haskell's default \ttp{Show} instances, which uses
record syntax.)  Functional programmers should be familiar with having to debug
a ``wall of text'' when debugging data structures.  What if instead a
counterexample like the following were returned by the testing framework:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
The formula quantifies away all the irrelevant portions of the data structure
with respect to the property, meaning that for any well-typed values that
replace \ttp{x0}, \ttp{x1}, \ttp{x2}, and \ttp{x3}, the counterexample still
holds.

@@

As demonstrated by a motivating example in Section~\ref{sec:example}, SmartCheck
can shrink more than an order-of-magnitude faster than efficient hand-written
shrink implementations for some properties and programs.  More importantly, the
time required to shrink scales linearly with the size of the
counterexample---other common shrink implementations often scale in polynomial
time or worse.  (Shrinking time is important in real-world uses; see
Section~\ref{sec:experiments} for a real-world example.)  And the
counterexamples returned are two to five times smaller.

@@

prop :: T -> Property
prop t = pre t ==> post t
\end{code}
  \caption{Example program and property.}
  \label{fig:initial}
\end{figure}

@@

A programmer might try to control the complexity by truncating lists using the
(\ttp{take n}) function that returns the first $n$ elements of a list.  The
trade-off is quicker shrinking with a lower probability of finding a smaller
counterexample.  For example, we might redefine shrink as follows:
%
\begin{samepage}
\begin{code}
shrink (T w i0 i1 i2 i3) =
  [ T a b c d e | a <- tk w
                , b <- tk i0, c <- tk i1
                , d <- tk i2, e <- tk i3 ]
  where tk x = take 10 (shrink x)
\end{code}
\end{samepage}
%
\noindent
Call this version ``truncated shrink''.  Truncation controls the blowup of the
input-space; the maximum number of possible new values is $10^5$ in this case.
While truncation controls the blowup, the downside is that potentially smaller
counterexamples may be omitted.

@@

SmallCheck is another testing framework for Haskell for which shrinking is
irrelevant.  SmallCheck is guaranteed to return a smallest counterexample, if
one exists~\cite{sc}.  SmallCheck does this by enumerating all possible inputs,
ordered from smallest to largest, up to some user-defined bound.  While
SmallCheck is effective for testing many programs and properties (in accordance
with the \emph{small scope hypothesis}~\cite{jackson}), counterexamples to even
relatively simple properties may be infeasible to discover due to input-space
explosion.

@@

For a product type, SmallCheck must check values produced by taking the
cross-product of the type's fields at a given depth.  Unfortunately, using
default instances, SmallCheck fails long before it reaches a depth required to
discover a counterexample.  (A related library named Feat combines some aspects
of SmallCheck and QuickCheck~\cite{feat}; we discuss it in
Section~\ref{sec:related}).

@@

@@

To motivate the benefits of SmartCheck reduction algorithm, consider the graphs
in Figure~\ref{fig:graphs} and the Table~\ref{table:results}, giving the results
for using QuickCheck and SmartCheck on the program shown in
Figure~\ref{fig:initial} (we omit SmallCheck, since as noted, it is not even
feasible for the example).  We graph counterexample size and execution time.
Out of 2000 runs, we show the number of runs resulting in a shrunk
counterexample of a given size (by counting \ttp{N} constructors), and we show
the number of runs executing within a given time limit (in seconds).  Note that
we provide two execution-time graphs, since the scale of time scales differ
dramatically.  In the table, we show the mean, median, median, and the results
at the 95th percentile values.  (The distributions produced by these values are
not necessarily Gaussian, so we omit the standard deviation.)

@@

More dramatic, however, are the differences in execution time.  The execution
times for SmartCheck include the initial counterexample generation time required
by QuickCheck; thus the time required by SmartCheck to shrink a counterexample
is on average just two hundredths of a second (0.10 - 0.08).  As can be seen in
the graphs, SmartCheck's execution time scales linearly (like QuickCheck,
without shrinking).  Truncation and tuple shrinking produce very long tails; the
longest execution time recorded for truncation shrinking and tuple shrinking are
88 seconds and 110 seconds, respectively.  SmartCheck is largely insensitive to
the size of the original counterexample provided.

The results for SmartCheck are fairly generalizable to other types and
properties since SmartCheck's reduction algorithm has no dependencies on the
specifics of either; however, user-defined shrink methods may vary significantly
based on the property and counterexample type.

@@

SmartCheck focuses on algebraic data types represented as ``sums of products''
including recursive and mutually recursive types.  We sometimes refer to the
elements of a sum type as a \emph{variant} that is \emph{tagged} by a
constructor.  For example, the type
%
\begin{code}
Maybe a = Just a | Nothing
\end{code}
%
\noindent
contains two variants, tagged by the constructors \ttp{Just} and \ttp{Nothing},
respectively.

To present the algorithms in the following sections, we provide some definitions
in the form of methods of a \ttp{SubTypes} type class.  (SmartCheck requires
instances to be defined for data being analyzed; in
Section~\ref{sec:implementation}, we describe how instances for the type class
are derived automatically.)
%
\begin{code}
type Idx    = Int
type Size   = Int
data SubVal = forall a. SubTypes a => SubVal a

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at an breadth-first index in a
  value,\footnote{Rather than indexes, an alternative representation is to use a
    zipper data structure~\cite{zipper}; see Section~\ref{sec:conclusions} for a
    discussion.}
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is true when the type of the value is an ``uninteresting
  type'', which generally means it is user-defined and has positive arity.
  We discuss the purpose of opaque types in Section~\ref{sec:base}.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

opaque (3 :: Int) = True
opaque tree       = False
opaque L          = False

@@

%% Finally, \ttp{opaque} is true for our tree type but false for a
%% structureless type, such as \ttp{Int}.

@@

The approach taken by typical \ttp{shrink} implementations is the most naive
graph search algorithm: it simply generates a random set of potential vertexes
that are strictly smaller according to the given measure to check in the graph.
However, it does not exploit that the original counterexample \ttp{cex} is
already in a neighborhood of other likely counterexamples.  So rather than
choose random vertexes, it may be more efficient to test other values that are
closely related to \ttp{cex}.  By restricting the search to \ttp{cex}'s
neighborhood, a smaller counterexample might be missed, and only a local minimum
is returned.  However, with a large set of values, finding a global minimum by
random sampling vertexes in the graph is possible but too inefficient, as we saw
from the example in the introduction.

\subsection{Reduction Algorithm Overview}\label{sec:reduct}
The algorithm we present for efficiently searching for new counterexamples is an
instance of greedy breadth-first search over a tree structure that represents a
value.  At each node, during the traversal, we generate arbitrary
\emph{structurally} smaller sub-values and build a new value from that, leaving
the remainder of the tree unchanged.  Intuitively, by a \emph{structurally
  smaller value}, we mean one with fewer constructors.  We continue until we
reach a fixed-point.

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  The reduction begins at the first
proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index \ttp{idx}
becomes out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.
Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => Size -> a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}: in the search space of possible values
from a given type \ttp{T}, if a known counterexample \ttp{cex} is a variant
\ttp{V} of \ttp{T}, then it is most probable that other counterexamples are also
from variant \ttp{V}.  Recall that the algorithm begins at the \emph{first} sub-value of
the counterexample rather than the 0th sub-value.  The 0th sub-value of
\ttp{cex} is \ttp{cex} itself.  No invariant of the algorithm would be violated
by beginning with the 0th sub-value, and in particular, the algorithm would
still terminate.  However, we begin with a strict sub-value because of the
hypothesis.  For example, consider a property about unintended variable capture
over a language's parse tree represented by a sum type with constructors for
module imports, function definitions, and global-variable assignments,
respectively.  A function definition counterexample can only be reduced to
smaller function definition counterexamples, the only construct in which
variable capture is possible.

@@

\subsubsection{Opaque Types}\label{sec:base}

The number of constructors in an algebraic data value primarily determines how
large it is.  In some cases, the size of fields that come from primitive
types, like \ttp{Int} or \ttp{Char}, have less of a role in understanding a
counterexample, so in general, we want to prevent our reduction algorithm from
spending time shrinking values from these types.

Consequently, we distinguish a class of types referred to as \emph{opaque
  types}.  If the reduction algorithm encounters a opaque type, it is ignored.
By default, this includes types without constructors (e.g., \ttp{Int}, \ttp{Word},
\ttp{Float}, \ttp{Bool}, and \ttp{Char}).  We also treat recursive container
types like lists specially: lists of opaque types are considered to be opaque
types themselves.  This prevents the algorithm from analyzing, for example,
strings.  A list of non-opaque-types, however, is shrunk by the algorithm.

By providing custom instances, the user can declare any substructure in a data
type to be a opaque type.  Doing so effectively treats values from that type as
``black boxes'' for SmartCheck.  Doing so can help make SmartCheck more
efficient if the user knows that some portion of the structure cannot be shrunk
or is irrelevant to the property.

Conversely, if the user wishes for types that are treated as opaque by default
by SmartCheck to be shrunk, the easiest approach is to define shrink instances
that operate on the opaque types while leaving the rest of the value constant.
These types are shrunk relatively efficiently by conventional QuickCheck's
pre-defined instances.

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}.  The
algorithm is similar to the reduction algorithm in Figure~\ref{fig:reduction},
and in the implementation, the algorithms are generalized and combined.  The
function \ttp{extrapolate} returns a list of indexes to be generalized in the
original counterexample.  In the recursive function \ttp{extrapolate'}, there is
a function guard in which a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
is true if in \ttp{cex}, the value at index \ttp{idx0} is a child of index
\ttp{idx1} in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  The
check needs only to compare indexes, so it is efficient.  The \ttp{subTrees}
guard prevents the algorithm from trying to generalize sub-values that are
abstracted away already since their parents have been generalized.

@@

\paragraph{Soundness}
The extrapolation algorithm is unsound in two ways.  First, it extrapolates from
a set of counterexamples to a universal claim, similar to QuickSpec or
Daikon~\cite{qs,daikon}.  By tuning the parameters, the risk of an unsound
generalization is reduced by requiring more or larger values to fail the
property.

Second, note that each index is generalized with respect to the concrete
sub-values that exist in the counterexample.  So in some cases, a formula may be
returned that is overly general.  For example, consider the counterexample in
which both fields of \ttp{Add} contain values causing the failure:
%
\begin{code}
Add (Div (C 1) (C (-2)) (C 2))
    (Div (C 0) (C (-1)) (C 1))
\end{code}
%
\noindent
since no matter what random value the first field of \ttp{Add} is replaced with,
the property fails by the second field and vice versa for the second.  The
universal generalization algorithm might return the formula
\noindent
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
An over-generalized formula is only a problem when generalizing without
minimizing the counterexample first, as pointed out to this author by John
Hughes, so in practice with SmartCheck, this form of unsoundness does not arise.
However, if it does, for universally quantified formulas, the reader should read
it as shorthand for quantifying each variable independently and taking the
conjunction of formulas.  For example, instead of
%
\begin{code}
forall values x0 x1 . Add x0 x1
\end{code}
%
\noindent
one should read
%
%
\begin{code}
      forall values x0 .
        Add x0 (Div (C 0) (C (-1)) (C 1))
  and forall values x1 .
        Add (Div (C 1) (C (-2)) (C 2)) x1
\end{code}

@@

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there is a
counterexample containing every possible variant as a sub-value, then we
abstract it.  For example, suppose that \ttp{prop\_div} had no precondition
\ttp{divSubTerms} testing for the occurrence of a constant (\ttp{C 0}) in the
divisor.  Then the following would all counterexamples:
%
\begin{code}
Div (Add (C 3) (C 2)) (C 0)
Div (C 7) (Add (C (-5)) (C 5))
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors c .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (c\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

Figure~\ref{fig:cex-gen} illustrates a property of the form (\ttp{pre ==>
  post}).  Points are specific counterexamples that satisfy the precondition but
fail the post-condition, and the enclosing ring represents the generalization of
counterexamples resulting from either universal or existential generalization.
Our goal is to find additional counterexamples in the shaded region.  As new
counterexamples are discovered in the shaded region (and generalized), the
counterexample space becomes covered until no more classes of counterexamples
exist or it becomes too difficult for the testing framework to discover them.

@@

Figure~\ref{fig:matches} shows the shape-matching algorithm used for precondition
strengthening.  The function takes a candidate counterexample and a list of
previous counterexamples, together with the indexes at which they have been
generalized.  The basic idea of the algorithm is to determine whether two values
have the same ``shape''.  Intuitively, we consider two values to have the same
shape if in their respective tree representations, their constructors at each
node in the tree match, ignoring all opaque types (Section~\ref{sec:base}).
That is, intuitively, for values \ttp{a} and \ttp{b},
%
\begin{code}
subVals a == subVals b
\end{code}
%
\noindent
Furthermore, indexes that have been universally generalized (and all their
children) match any value, since a universally generalized index indicates that
counterexamples have been found for any possible constructor.  In our
implementation, we also consider existentially generalized indexes to match any
value.  Doing so is a design choice that is more aggressive about covering the
space of counterexamples at the risk of omitting some.
